
%{\small
\begin{thebibliography}{9}

\bibitem{ahoullman} 
   A.~V.~Aho, J.~D.~Ullman. \emph{Foundations of Computer Science}, 
Computer Science Press, 1992.

\bibitem{bentley-prog-pearls}
J. Bentley. \emph{Programming Pearls}, Second Edition. 
Addison-Wesley, Inc., 2000.

\bibitem{brandenburg} 
F.~J.~ Brandenburg. \emph{The Graph Template Library---GTL},\\ 
(see \url{http://www.infosun.fim.uni-passau.de/GTL/})

\bibitem{cormen} 
T.~H.~ Cormen, C.~E.~Leiserson and R.~L.~Rivest.
      \emph{Introduction to Algorithms}, McGraw-Hill, New York, 1990. 

\bibitem{Ed65}
J. Edmonds.
``Paths, trees, and flowers,''
\emph{Canadian Journal of Mathematics} 17 (1965), pages 449--467. 

\bibitem{flaj-sedg}
R.~Sedgewick and P.~Flajolet.
\emph{Introduction to the Analysis of Algorithms}, Addison-Wesley, Inc., 1996.

\bibitem{FF56}
L. R. Ford and D. R. Fulkerson 
``Maximal flow through a network,"
\textit{Canadian Journal of Mathematics} \textbf{8} (1956), pages
399--404.

\bibitem{goodrichtamassia}
M.~T.~ Goodrich and R.~ Tamassia. 
       \emph{Data Structures and Algorithms in Java}, 
       John Wiley and Sons, Inc., 2001.

\bibitem{mehlhorn}
K.~ Mehlhorn and St.~ Nadher.
   \emph{The LEDA Platform of Combinatorial and Geometric Computing},
   Cambridge University Press, 1999. \\
   (see \url{http://www.mpi-sb.mpg.de/LEDA/leda.html})

\bibitem{orwant}
   J.~ Orwant, J.~ Hietaniemi and J.~ Macdonald.
    \emph{Mastering Algorithms with Perl}, O'Reilly, August 1999.

\bibitem{Boost}
    J.G. Siek, L-Q. Lee and A. Lumsdaine.
\emph{The Boost Graph Library: User Guide and Reference Manual}
    Addison-Wesley, 2001. (see \url{http://www.boost.org}) 

\bibitem{standish} 
   T.~A.~Standish. \emph{Data Structures in Java\(^{\rm TM}\)}, 
                 Addison-Wesley, 1998.

\end{thebibliography} 
%}
