%\setcounter{chapter}{-1}
\chapter*{Introduction to the Second Edition}
\markboth{Introduction to the Second Edition}{Introduction to the Second Edition}
%\pagestyle{myheadings}

Writing a second edition is a thankless task, as is well known to authors. Much 
of the time is spent on small improvements that are not obvious to readers. 
We have taken considerable efforts to correct a large number of errors found 
in the first edition, and to improve explanation and presentation throughout the 
book, while retaining the philosophy behind the original. As far as material 
goes, the main changes are:
\begin{itemize}\setlength{\itemsep}{-1pt}
\item more exercises and solutions to many of them;
\item a new section on maximum matching (Section~\ref{sec:matching});
\item a new section on string searching (Part III); % (Section~\ref{sec:strmatch});
\item a Java graph library updated to Java 1.6 and freely available for download.
\end{itemize} 

The web site \url{http://www.cs.auckland.ac.nz/textbookCS220/} for the book
provides additional material including source code. Readers finding errors are
encouraged to contact us after viewing the errata page at this web site.

In addition to the acknowledgments in the first edition, we thank
Sonny Datt for help with updating the Java graph library,
Andrew Hay for help with exercise solutions and Cris Calude for comments.
Rob Randtoul (\texttt{PlasmaDesign.co.uk}) kindly allowed 
us to use his cube artwork for the book's cover.
Finally, we thank

\begin{description}
\item [MJD] all students who have struggled to learn from the first edition and 
have given us feedback, either positive or negative; 

\item [GLG] my wife Natasha and all the family for their permanent help 
and support;

\item[MCW] my wife Golbon and sons Yusef and Yahya, for their sacrifices 
during the writing of this book, and the joy they bring to my life even in 
the toughest times.
\end{description}

\bigskip

\begin{flushright}
\if 01
Michael J. Dinneen\\
Georgy Gimel'farb\\
Mark C. Wilson\\[2ex]

Department of Computer Science\\ 
University of Auckland\\
Auckland, New Zealand\\[2ex]

\verb|{mjd,georgy,mcw}@cs.auckland.ac.nz|\\[2ex]
\fi
31 October 2008
\end{flushright}




